\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{booktabs}
\usepackage{array}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage[table]{xcolor}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue,
  citecolor=blue,
  pdftitle={HSA as Sparse Causal Routing and a Path Toward Typed Semantic Graphs},
  pdfauthor={OpenAI Codex}
}

\newcolumntype{C}{>{\centering\arraybackslash}m{1.5em}}
\newcommand{\mempty}{\cellcolor{gray!10}\rule{0pt}{1.5ex}\hspace{0.75em}}
\newcommand{\mlive}{\cellcolor{blue!20}\rule{0pt}{1.5ex}\hspace{0.75em}}
\newcommand{\mrowpick}{\cellcolor{green!25}\rule{0pt}{1.5ex}\hspace{0.75em}}
\newcommand{\mcolpick}{\cellcolor{violet!22}\rule{0pt}{1.5ex}\hspace{0.75em}}
\newcommand{\mcore}{\cellcolor{green!30}\rule{0pt}{1.5ex}\hspace{0.75em}}
\newcommand{\mtail}{\cellcolor{orange!30}\rule{0pt}{1.5ex}\hspace{0.75em}}

\title{HSA as Sparse Causal Routing\\
\large Formalization, Current Packing Strategy, and a Path Toward Typed Semantic Graphs}
\author{}
\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
This note formalizes the current hierarchical sparse attention (HSA) stack as a
decoder-causal sparse routing problem over a sequence graph, explains the
current dense HSA reference, schedule-based sparse realization, and packing
strategy, and records why \texttt{sparse\_mask} backward remains the pragmatic
training path even though it uses a coarser sparsification than the most
aggressive forward-only packed routes. The note then sketches a realistic
``leapfrog'' extension path from binary hierarchical masks toward typed,
directed, and eventually signed semantic routing graphs that could support
precursors to causal reasoning in autoregressive language models. The central
claim is that the current implementation is not ``arbitrary graph attention,''
but it already contains reusable pieces of a graph compiler, sparse runtime, and
offline sidecar system that can be repurposed for richer semantic graphs.
\end{abstract}

\section{Motivation}

The current HSA recipe uses semantic structure already present in text
representation---sentence anchors, section anchors, document anchors, or other
hierarchical boundary markers---to build a sparse causal attention graph. This
graph is then realized either as a dense semantic mask or as an equivalent
schedule-based sparse computation.

The long-range motivation is twofold:
\begin{enumerate}[leftmargin=1.5em]
    \item reduce attention cost while preserving relevance over long contexts,
    \item bias the model toward structured information flow rather than letting
    dense attention allocate probability mass over all causal predecessors.
\end{enumerate}

That is already useful for ``context rot'' mitigation, but it also opens a more
ambitious direction: moving from document hierarchy to typed semantic graphs and
eventually to typed directional message passing that can approximate parts of
causal reasoning without claiming true interventional semantics.

\section{Formalizing Current HSA}

\subsection{Sequence, graph, and support mask}

Let a decoder consume tokens $x_1, \dots, x_T$. Standard causal self-attention
defines a dense lower-triangular support:
\[
\mathcal{E}_{\mathrm{causal}} = \{(i,j) \mid 1 \le j \le i \le T\}.
\]

Current HSA replaces this with a sparse causal subgraph
\[
G = (V, E), \quad V = \{1, \dots, T\}, \quad E \subseteq \mathcal{E}_{\mathrm{causal}}.
\]

Equivalently, it defines a binary support mask
\[
S_{ij} =
\begin{cases}
1 & \text{if } (i,j) \in E,\\
0 & \text{otherwise.}
\end{cases}
\]

The graph $E$ is not learned online. It is compiled from token-side semantic
metadata such as \texttt{keep\_ids} and \texttt{hash\_ids}, which encode the
hierarchical structure of the document.

\subsection{Dense HSA reference}

Current dense HSA is a masked exact attention reference:
\[
\mathrm{score}_{ij} = \frac{q_i^\top k_j}{\sqrt{d}} + m_{ij},
\]
where
\[
m_{ij} =
\begin{cases}
0 & \text{if } S_{ij}=1,\\
-\infty & \text{if } S_{ij}=0.
\end{cases}
\]
Then
\[
\alpha_{ij} = \frac{\exp(\mathrm{score}_{ij})}
{\sum_{\ell : S_{i\ell}=1}\exp(\mathrm{score}_{i\ell})},
\qquad
o_i = \sum_{j : S_{ij}=1} \alpha_{ij} v_j.
\]

This is the clean semantic reference: all questions about correctness reduce to
whether the dense mask implements the intended graph.

\subsection{Schedule-based sparse HSA}

The schedule-based sparse implementation is supposed to compute exactly the same
function as the dense mask, but using explicit sparse descriptors rather than a
dense masked traversal. In the repository, that compiler/runtime split lives
primarily in:
\begin{itemize}[leftmargin=1.5em]
    \item \texttt{flash\_attn/cute/hsa.py},
    \item \texttt{flash\_attn/cute/hsa\_cached\_2d\_forward\_analysis.py},
    \item \texttt{nanochat/hsa\_batch\_cache.py}.
\end{itemize}

The schedule is therefore a \emph{compiled sparse realization of the same
binary support graph}, not a different model class.

\section{What the Current Graph Semantics Actually Encode}

Current HSA is not general graph attention. It is a decoder-causal sparse graph
with a specific information-routing policy derived from hierarchy.

In the present recipe, the graph includes:
\begin{itemize}[leftmargin=1.5em]
    \item same-sentence body-level connectivity,
    \item same-section anchor connectivity,
    \item same-document anchor connectivity,
    \item cross-level edges such as body $\rightarrow$ previous sentence anchor,
    \item body $\rightarrow$ previous section anchor,
    \item sentence anchor $\rightarrow$ previous section anchor.
\end{itemize}

These cross-level edges matter. Without them, the graph is too lossy: a purely
same-level tree-shaped routing policy under-connects the model and discards
useful long-range semantic pathways.

The important subtlety is that this graph is a \emph{routing graph}, not an
ontology. The document hierarchy is a tree. Useful attention flow is closer to a
causalized directed acyclic graph over the prefix.

\section{Dense HSA Versus Sparse HSA}

\subsection{Dense HSA}

Dense HSA means: build the exact semantic mask, run the ordinary dense masked
attention reference, and let the kernel traverse all entries in the lower
triangle even if many entries are masked out. This is conceptually simple and is
the best semantic oracle when debugging graph meaning.

\subsection{Sparse HSA}

Sparse HSA means: compile the same semantic mask into sparse descriptors and
realize the same attention via sparse traversal, sparse runtime metadata, or
packed subproblems. Sparse HSA is therefore a systems optimization problem on
top of the same graph semantics.

\subsection{Invariant that matters}

The critical invariant is:
\[
\text{dense semantic mask} \equiv \text{sparse schedule} \equiv \text{kernel support}.
\]

When these differ, training behavior diverges even if the intended high-level
semantics were correct. Much of the engineering difficulty in HSA is therefore
not inventing a graph, but proving that every layer of the implementation
realizes the same graph.

\section{Why Packing Helps Forward More Than Backward}

\subsection{Forward is naturally query-owned}

Forward attention on a compressed subproblem is naturally row-owned:
\[
Q_b \in \mathbb{R}^{r \times d}, \quad
K_b, V_b \in \mathbb{R}^{u \times d},
\]
with one output $O_b$ and one log-sum-exp state per query row. Packing reduces
an irregular sparse neighborhood into a regular dense microproblem.

That is why forward benefits strongly from packing:
\begin{itemize}[leftmargin=1.5em]
    \item gather and pack once,
    \item compute one tiny dense exact attention,
    \item write one output and one LSE per row.
\end{itemize}

\subsection{Backward has an ownership mismatch}

Backward is harder because the natural ownership splits:
\[
dQ = dS K \quad \text{(query-row owned)}
\]
but
\[
dK \mathrel{+}= dS^\top Q, \qquad
dV \mathrel{+}= P^\top dO
\quad \text{(key-row owned)}.
\]

This means the same forward packing that makes a tiny dense forward problem does
not automatically make a tiny dense backward problem with a natural writeback
owner. The moment we leave the query-owned world, we pay for reductions,
atomics, or extra saved-state movement.

\subsection{Why \texttt{sparse\_mask} backward is still so good}

Even though \texttt{sparse\_mask} backward is coarser and often less elegant
than the most compressed forward path, it remains strong for several reasons:
\begin{enumerate}[leftmargin=1.5em]
    \item it stays close to the exact support graph;
    \item it amortizes work in large regular backward kernels instead of carrying
    fragile packed forward state into backward;
    \item it avoids the q-owned / k-owned mismatch becoming a bespoke packed
    writeback problem;
    \item once numerical issues are fixed, it is robust and easy to validate
    against the dense masked reference;
    \item empirically, the forward gains from aggressive packing are large,
    while the backward gains from the same packing are usually small or negative.
\end{enumerate}

This is the main reason the hybrid path---compressed forward plus
\texttt{sparse\_mask} backward---has remained the best practical training
configuration.

\section{Row Packing, Column Packing, and Generalized 2-D Packing}

\subsection{Row packing}

Row packing compresses multiple query rows with similar neighborhoods into a
single regular microproblem. Informally:
\begin{itemize}[leftmargin=1.5em]
    \item choose a small set of query rows,
    \item take the union of their live keys,
    \item gather that union once,
    \item run one tiny dense attention over the packed rows.
\end{itemize}

This is the most natural packing because forward attention is query-owned.

\subsection{Column packing}

Column packing focuses on the key/value side:
\begin{itemize}[leftmargin=1.5em]
    \item compress repeated or shared key neighborhoods,
    \item reduce redundant K/V gathers,
    \item try to reuse union-K structure across rows.
\end{itemize}

This helps when many rows share similar key support, but on its own it is less
complete than row packing because it does not resolve the query-side ownership
pattern.

\subsection{Generalized 2-D packing}

Generalized 2-D packing treats the support matrix as a sparse binary matrix and
tries to extract dense submatrices:
\begin{itemize}[leftmargin=1.5em]
    \item exact dense tiles when a row group shares a common live key set,
    \item tiled residual tails for uncovered support,
    \item range-level fused execution where exact tiles and tails share one
    online softmax state.
\end{itemize}

That is more general than either pure row packing or pure column packing:
\begin{itemize}[leftmargin=1.5em]
    \item row packing asks ``which rows should be grouped?''
    \item column packing asks ``which K/V neighborhoods should be reused?''
    \item generalized 2-D packing asks ``which dense submatrices can be
    extracted from the sparse support matrix, and how should the remainder be
    tiled?''
\end{itemize}

The current cached generalized forward path in the repository is exactly this
kind of 2-D extraction:
\begin{itemize}[leftmargin=1.5em]
    \item exact dense core tiles,
    \item fused tiled tail,
    \item one range-local softmax state,
    \item one output write per row range.
\end{itemize}

\subsection{Packing illustrations}

Figure~\ref{fig:packing-illustrations} shows the three packing viewpoints on a
small support matrix. Blue cells are live edges in the original sparse support.
Green cells indicate an extracted dense core, purple cells indicate a reused
column family, and orange cells indicate a tiled residual tail.

\begin{figure}[ht]
\centering
\setlength{\tabcolsep}{2pt}
\renewcommand{\arraystretch}{1.1}

\begin{minipage}[t]{0.31\textwidth}
\centering
\textbf{Row packing}\\[0.35em]
\begin{tabular}{c|*{6}{C}}
 & $k_1$ & $k_2$ & $k_3$ & $k_4$ & $k_5$ & $k_6$ \\
\hline
$q_1$ & \mrowpick & \mrowpick & \mempty & \mrowpick & \mempty & \mempty \\
$q_2$ & \mrowpick & \mrowpick & \mempty & \mrowpick & \mempty & \mempty \\
$q_3$ & \mempty & \mempty & \mlive & \mempty & \mlive & \mempty \\
$q_4$ & \mempty & \mempty & \mlive & \mempty & \mlive & \mlive \\
\end{tabular}

\vspace{0.5em}

\begin{tabular}{c|*{3}{C}}
 & $u_1$ & $u_2$ & $u_3$ \\
\hline
$q_1$ & \mcore & \mcore & \mcore \\
$q_2$ & \mcore & \mcore & \mcore \\
\end{tabular}

\vspace{0.35em}
\raggedright
\footnotesize
Select similar rows $\{q_1,q_2\}$, form the union
$\{k_1,k_2,k_4\}$ once, and run one dense $2 \times 3$ microproblem.
\end{minipage}
\hfill
\begin{minipage}[t]{0.31\textwidth}
\centering
\textbf{Column packing}\\[0.35em]
\begin{tabular}{c|*{6}{C}}
 & $k_1$ & $k_2$ & $k_3$ & $k_4$ & $k_5$ & $k_6$ \\
\hline
$q_1$ & \mcolpick & \mcolpick & \mempty & \mcolpick & \mempty & \mempty \\
$q_2$ & \mcolpick & \mcolpick & \mempty & \mcolpick & \mempty & \mempty \\
$q_3$ & \mempty & \mempty & \mlive & \mempty & \mlive & \mempty \\
$q_4$ & \mempty & \mempty & \mlive & \mempty & \mlive & \mlive \\
\end{tabular}

\vspace{0.5em}

\begin{tabular}{c|*{3}{C}}
 & $k_1$ & $k_2$ & $k_4$ \\
\hline
reuse & \mcolpick & \mcolpick & \mcolpick \\
\end{tabular}

\vspace{0.35em}
\raggedright
\footnotesize
Identify a repeated K/V family and gather those columns once. This reduces
redundant K/V movement, but row ownership is still unresolved by itself.
\end{minipage}
\hfill
\begin{minipage}[t]{0.31\textwidth}
\centering
\textbf{Generalized 2-D packing}\\[0.35em]
\begin{tabular}{c|*{6}{C}}
 & $k_1$ & $k_2$ & $k_3$ & $k_4$ & $k_5$ & $k_6$ \\
\hline
$q_1$ & \mcore & \mcore & \mempty & \mcore & \mempty & \mempty \\
$q_2$ & \mcore & \mcore & \mempty & \mcore & \mempty & \mempty \\
$q_3$ & \mempty & \mempty & \mtail & \mempty & \mtail & \mempty \\
$q_4$ & \mempty & \mempty & \mtail & \mempty & \mtail & \mtail \\
\end{tabular}

\vspace{0.5em}

\begin{tabular}{c|*{3}{C}}
 & $u_1$ & $u_2$ & $u_3$ \\
\hline
core & \mcore & \mcore & \mcore \\
tail & \mtail & \mtail & \mtail \\
\end{tabular}

\vspace{0.35em}
\raggedright
\footnotesize
Extract dense submatrices from the support itself: an exact dense core tile for
$(q_1,q_2)$, then cover the remainder with tiled residual tails that share one
range-local softmax state.
\end{minipage}

\caption{Three views of packing. Row packing groups similar query rows, column
packing reuses repeated K/V families, and generalized 2-D packing extracts
dense submatrices from the support matrix and covers the remainder with fused
tail tiles.}
\label{fig:packing-illustrations}
\end{figure}

The important systems point is that all three methods operate on the support
graph rather than on the exact scoring rule. This means the graph compiler,
sidecars, schedule families, and most of the current packing machinery remain
reusable if the score becomes relation-aware or typed. The part that changes is
the kernel math inside each packed family: score evaluation, value transforms,
and possibly normalization if the model moves beyond plain masked softmax.

\section{What the Current Implementation Already Gives Us}

The current stack contains several reusable components that are more general
than ``hierarchy only'' suggests.

\subsection{1. A graph compiler from token annotations to sparse causal support}

Today this compiler is expressed through:
\begin{itemize}[leftmargin=1.5em]
    \item \texttt{keep\_ids},
    \item \texttt{hash\_ids},
    \item \texttt{compute\_hsa\_mask(...)},
    \item \texttt{build\_hsa\_schedule(...)}.
\end{itemize}

This is already a graph compiler. It just currently compiles one specific
family of hierarchical binary support graphs.

\subsection{2. Exact dense and sparse references}

The repository already has the conceptual machinery needed to separate:
\begin{itemize}[leftmargin=1.5em]
    \item intended semantics,
    \item sparse schedule,
    \item kernel behavior.
\end{itemize}

That is crucial for any richer graph scheme. If the graph is extended to typed
or directed edges, the same discipline will still be necessary.

\subsection{3. Sidecars and offline precomputation}

The schedule sidecar and forward sidecar system is not specific to hierarchy.
It is a reusable vehicle for:
\begin{itemize}[leftmargin=1.5em]
    \item offline graph compilation,
    \item offline packing,
    \item runtime attachment of structural metadata,
    \item avoiding host-side rebuild costs.
\end{itemize}

This becomes even more valuable if the graph moves from simple hierarchy to
typed semantic edges, since graph compilation will likely become more expensive.

\subsection{4. Packing policy and schedule families}

The current row-compact and generalized 2-D packing framework is a reusable
``graph-to-kernel-family'' compiler. The main idea is already in place:
\begin{itemize}[leftmargin=1.5em]
    \item start from an irregular support graph,
    \item partition it into a small set of regular families,
    \item run specialized kernels on each family.
\end{itemize}

That is exactly the right systems pattern for richer semantic graphs too.

\section{What Does \emph{Not} Recycle Directly}

Not every part of the current HSA stack is directly reusable for a leap toward
semantic or causal graphs.

\subsection{1. Binary support is too weak for typed semantics}

If all edges are reduced to a single binary mask, then
\[
\texttt{Spain} \to \texttt{Europe}
\quad\text{and}\quad
\texttt{rain} \to \texttt{wet}
\]
look identical at the routing level. The graph topology survives, but relation
type does not.

\subsection{2. Standard softmax attention only models positive mixing weights}

Softmax produces
\[
\alpha_{ij} \ge 0, \qquad \sum_j \alpha_{ij} = 1.
\]
This is appropriate for compatibility-based readout, but it is not a natural
model of signed causal influence. The output can still become negative through
the value vectors, but the weighting mechanism itself is not signed.

\subsection{3. Current kernels assume binary support plus standard reductions}

The current sparse kernels assume:
\begin{itemize}[leftmargin=1.5em]
    \item allowed or forbidden edges,
    \item standard score computation,
    \item standard softmax normalization,
    \item standard $QK^\top$ and weighted $V$ reductions.
\end{itemize}

Typed, signed, or relation-conditioned message passing would require either:
\begin{itemize}[leftmargin=1.5em]
    \item multiple edge-family kernels,
    \item relation-specific score transforms,
    \item relation-specific value transforms,
    \item or an alternative aggregator that is no longer a pure masked softmax.
\end{itemize}

\section{From Hierarchy to Typed Directed Semantic Graphs}

\subsection{The right abstraction}

The natural extension is not ``arbitrary graph attention'' in the abstract. The
natural extension is:
\begin{quote}
decoder-causal sparse routing over tokens and semantic memory nodes, with typed
directed edges.
\end{quote}

This still respects autoregressive training, but allows the graph to encode
more than document boundaries.

\subsection{Semantic direction versus attention direction}

There are two distinct notions of direction:
\begin{enumerate}[leftmargin=1.5em]
    \item \textbf{semantic direction:} e.g.\ a knowledge or event relation,
    \item \textbf{attention-flow direction:} which previously visible node a
    token is allowed to read from.
\end{enumerate}

These are not the same thing. A semantic graph may contain a relation
\[
\texttt{rain} \to \texttt{wet},
\]
but in a decoder we still must causalize it into past-visible reads only. So a
semantic graph must be compiled into a causal routing graph before it becomes an
attention support.

\subsection{A minimal typed extension}

The smallest meaningful extension from current HSA is:
\begin{itemize}[leftmargin=1.5em]
    \item keep binary support families,
    \item add edge type $\tau_{ij}$,
    \item modify the score and/or transmitted message by type.
\end{itemize}

A minimal relation-aware score is:
\[
\mathrm{score}_{ij}
=
q_i^\top W_{\tau_{ij}} k_j + b_{\tau_{ij}}.
\]

A minimal relation-aware message is:
\[
m_{ij}
=
\alpha_{ij}\, T_{\tau_{ij}}(v_j),
\qquad
o_i = \sum_{j} m_{ij}.
\]

This already gives a typed directed semantic graph while remaining relatively
close to the current masked-attention implementation model.

\section{Toward Signed and Causal Message Passing}

\subsection{Why signed edges are interesting}

Some relations are naturally excitatory while others are inhibitory. For
example, a world-state graph might treat
\begin{itemize}[leftmargin=1.5em]
    \item \texttt{rain} $\to$ \texttt{wet-ground} as positive,
    \item \texttt{watering} $\to$ \texttt{wet-ground} as positive,
    \item \texttt{sun/heat} $\to$ \texttt{wet-ground} as negative or drying.
\end{itemize}

A pure binary mask cannot distinguish these. Even a typed score only changes who
is read from, not whether the message should reinforce or inhibit the target.

\subsection{What softmax cannot express directly}

Softmax gives positive simplex weights. So if the goal is to model signed or
inhibitory influence directly, one likely needs one of:
\begin{enumerate}[leftmargin=1.5em]
    \item relation-specific value transforms with sign,
    \item dual positive/negative channels,
    \item a non-simplex aggregator,
    \item latent semantic state nodes where inhibitory structure is carried in
    state rather than in the raw attention weights.
\end{enumerate}

\subsection{A realistic near-term precursor}

The most practical precursor to causal reasoning is not true interventional
causality. It is:
\begin{quote}
typed directed sparse message passing over tokens and semantic memory nodes,
where the graph encodes permitted information pathways and relation-specific
message transforms.
\end{quote}

That is already a meaningful leap beyond hierarchy, while remaining compatible
with autoregressive training.

\section{Why This Is Not Yet True Causal Reasoning}

A directed semantic graph inside a next-token predictor is still an
\emph{inductive bias}, not a causal model in Pearl's sense. True causal claims
require more than directional edges:
\begin{itemize}[leftmargin=1.5em]
    \item intervention semantics,
    \item counterfactual structure,
    \item assumptions about latent confounding,
    \item or explicit world-state modeling.
\end{itemize}

So the better description is:
\begin{quote}
the current HSA machinery can be extended into a typed semantic routing
framework that may support precursors to causal reasoning, but not causal
inference merely by virtue of using a directed graph.
\end{quote}

\section{A Concrete Leapfrog Plan}

\subsection{Stage 1: typed sparse graph}

Extend the current binary support into a small set of typed edge families:
\begin{itemize}[leftmargin=1.5em]
    \item local,
    \item summary/hierarchy,
    \item entity-link,
    \item discourse,
    \item causal-positive,
    \item causal-negative.
\end{itemize}

This step can reuse:
\begin{itemize}[leftmargin=1.5em]
    \item token-side metadata extraction,
    \item graph compilation,
    \item sidecars,
    \item packed family scheduling.
\end{itemize}

\subsection{Stage 2: relation-aware score and message transforms}

Keep the sparse routing framework, but make score and message depend on edge
type. This is still close enough to current HSA that the implementation can
evolve incrementally.

\subsection{Stage 3: semantic memory nodes}

Introduce explicit non-token nodes such as:
\begin{itemize}[leftmargin=1.5em]
    \item entity nodes,
    \item event nodes,
    \item state nodes,
    \item section or chapter summaries,
    \item retrieved knowledge nodes.
\end{itemize}

Then sparse attention becomes routing between tokens and memory/state nodes,
rather than only between raw tokens.

\subsection{Stage 4: signed or non-simplex aggregation}

Only after typed graph routing is stable should the scoring rule itself be
generalized beyond pure masked softmax, for example by allowing signed message
coefficients or separate positive/negative channels.

\section{Summary}

The current HSA implementation should be understood as \emph{sparse causal
routing}, not arbitrary graph attention. Its graph compiler, exact dense/sparse
references, sidecar system, and packing framework already provide reusable
infrastructure for a richer semantic-graph program.

The key engineering lessons so far are:
\begin{enumerate}[leftmargin=1.5em]
    \item correctness depends on semantic mask, sparse schedule, and kernel
    support all matching exactly;
    \item forward benefits strongly from aggressive packing because the problem
    is query-owned and easy to regularize;
    \item backward remains best as \texttt{sparse\_mask} because the q-owned /
    k-owned mismatch makes packed backward substantially harder;
    \item generalized 2-D packing is the right forward abstraction because it
    extracts dense submatrices from the support matrix itself;
    \item the natural next leap is typed directed semantic routing, not a jump
    straight to unrestricted signed graph inference.
\end{enumerate}

In short: the current method will not work with \emph{any} graph and token
scheme, but it is already close to a general graph compiler for \emph{causalized
sparse semantic routing}. That is a strong base for the next stage.

\section*{References}

\begin{thebibliography}{99}

\bibitem{vaswani2017attention}
Ashish Vaswani, Noam Shazeer, Niki Parmar, Jakob Uszkoreit, Llion Jones,
Aidan N. Gomez, Lukasz Kaiser, and Illia Polosukhin.
\newblock \emph{Attention Is All You Need}.
\newblock NeurIPS, 2017.
\newblock \url{https://arxiv.org/abs/1706.03762}

\bibitem{dao2022flashattention}
Tri Dao, Daniel Y. Fu, Stefano Ermon, Atri Rudra, and Christopher Re.
\newblock \emph{FlashAttention: Fast and Memory-Efficient Exact Attention with IO-Awareness}.
\newblock NeurIPS, 2022.
\newblock \url{https://arxiv.org/abs/2205.14135}

\bibitem{dao2023flashattention2}
Tri Dao.
\newblock \emph{FlashAttention-2: Faster Attention with Better Parallelism and Work Partitioning}.
\newblock 2023.
\newblock \url{https://arxiv.org/abs/2307.08691}

\bibitem{beltagy2020longformer}
Iz Beltagy, Matthew E. Peters, and Arman Cohan.
\newblock \emph{Longformer: The Long-Document Transformer}.
\newblock 2020.
\newblock \url{https://arxiv.org/abs/2004.05150}

\bibitem{zaheer2020bigbird}
Manzil Zaheer, Guru Guruganesh, Kumar Avinava Dubey, Joshua Ainslie,
Chris Alberti, Santiago Ontanon, Philip Pham, Anirudh Ravula,
Qifan Wang, Li Yang, and Amr Ahmed.
\newblock \emph{Big Bird: Transformers for Longer Sequences}.
\newblock NeurIPS, 2020.
\newblock \url{https://arxiv.org/abs/2007.14062}

\bibitem{velivckovic2018gat}
Petar Velickovic, Guillem Cucurull, Arantxa Casanova, Adriana Romero,
Pietro Lio, and Yoshua Bengio.
\newblock \emph{Graph Attention Networks}.
\newblock ICLR, 2018.
\newblock \url{https://arxiv.org/abs/1710.10903}

\bibitem{schlichtkrull2018rgcn}
Michael Schlichtkrull, Thomas N. Kipf, Peter Bloem, Rianne van den Berg,
Ivan Titov, and Max Welling.
\newblock \emph{Modeling Relational Data with Graph Convolutional Networks}.
\newblock ESWC, 2018.
\newblock \url{https://arxiv.org/abs/1703.06103}

\bibitem{battaglia2018relational}
Peter W. Battaglia, Jessica B. Hamrick, Victor Bapst, Alvaro Sanchez-Gonzalez,
Vinicius Zambaldi, Mateusz Malinowski, Andrea Tacchetti, David Raposo,
Adam Santoro, Ryan Faulkner, Caglar Gulcehre, H. Francis Song,
Andrew J. Ballard, Justin Gilmer, George E. Dahl, Ashish Vaswani,
Kelsey R. Allen, Charles Nash, Victoria Langston, Chris Dyer,
Nicolas Heess, Daan Wierstra, Pushmeet Kohli, Matthew Botvinick,
Oriol Vinyals, Yujia Li, and Razvan Pascanu.
\newblock \emph{Relational Inductive Biases, Deep Learning, and Graph Networks}.
\newblock 2018.
\newblock \url{https://arxiv.org/abs/1806.01261}

\bibitem{hu2020hgt}
Ziniu Hu, Yuxiao Dong, Kuansan Wang, and Yizhou Sun.
\newblock \emph{Heterogeneous Graph Transformer}.
\newblock WWW, 2020.
\newblock \url{https://arxiv.org/abs/2003.01332}

\bibitem{derr2018signedgcn}
Tyler Derr, Yao Ma, and Jiliang Tang.
\newblock \emph{Signed Graph Convolutional Network}.
\newblock ICDM, 2018.
\newblock \url{https://arxiv.org/abs/1808.06354}

\bibitem{pearl2009causality}
Judea Pearl.
\newblock \emph{Causality: Models, Reasoning, and Inference}.
\newblock Cambridge University Press, second edition, 2009.
\newblock \url{https://www.cambridge.org/core/books/causality/B0046844FAE10CBF274D4ACBDAEB5F5B}

\end{thebibliography}

\end{document}
